Parallel Adaptive Importance Sampling

### 2017-01-22 PAIS.tex
changes to numerics section - removal of B2 and Chem problem

\documentclass[final]{siamltex}
%test change
\usepackage{cite}
\usepackage{graphicx,bbm,pstricks}
\usepackage{pifont}
\usepackage{bbm,algorithmic,mdframed,placeins,multirow,booktabs,subfigure}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}

@@

\def\leftrightarrowfill@{%
 \arrowfill@\leftarrow\relbar\rightarrow%
 }  
\makeatother

@@

\begin{table}[!h]
\centering
\begin{algorithm}[H]
\DontPrintSemicolon
\BlankLine
	Set $\X^{(0)} = \X_0 = [x_1^{(0)},x_2^{(0)},\ldots,x_M^{(0)}]^\top$.\;
	\For {$i = 1, \dots, N$}
	{
		Sample $\Y^{(i)} = [y_1^{(i)},y_2^{(i)},\ldots,y_M^{(i)}]^\top, \quad y_j^{(i)} \sim
\nu(\cdot;x_j^{(i-1)})$.\label{algline:PAIS_propose}\;
		Calculate $\W^{(i)} = [w_1^{(i)},w_2^{(i)},\ldots,w_M^{(i)}]^\top,$ \quad $w^{(i)}_j =
\frac{\pi(y_j^{(i)})}{\chi(y_j^{(i)};\X^{(i-1)})}$, where
		\[
			\chi(\cdot;\X^{(i-1)}) = \frac{1}{M}\sum_{j=1}^M \nu(\cdot;x_j^{(i-1)}).
		\]

@@

\section{Automated tuning of Algorithm Parameters}\label{Sec:adapt}

@@

\begin{table}[!h]
\centering
\begin{algorithm}[H]
\DontPrintSemicolon
\BlankLine
	Define update times $\{n_k\}_k$.\;
	\For{$n=1,\dots,N$}{
		Complete steps \ref{algline:PAIS_propose}-\ref{algline:PAIS_resample} of Algorithm~\ref{alg:PAIS}.\;
		\If {$n \in \{n_k\}$}
		{
			Divide ensemble into two halves. Use these halves to estimate the gradient in $\neff$ at $\delta_k$.\;
			Update $\delta_k$ using gradient ascent,
				\[
					\delta_{k+1} = \delta_k + \gamma\nabla\neff.
				\]
			\label{algline:gradient_ascent}
		}
	}
\caption{Adaptively tuned PAIS algorithm.\label{alg:adaptPAIS}}
\end{algorithm}
\end{table}
Our adaptive algorithm is given in Algorithm~\ref{alg:adaptPAIS}. We choose update times which, as suggested in Section~\ref{sec:ess} allow for a reasonable estimate of the effective sample size, but do not waste too many iterations. In Step~\ref{algline:gradient_ascent}, the gradient ascent parameter $\gamma$ may decrease over time, e.g. as a function of $n_k$.

@@

\begin{table}[!h]
\centering
\begin{algorithm}[H]
\DontPrintSemicolon
\BlankLine
	$\b{z} = M\bar{\b{w}}$.\;
	\For {$i = 1,\dots, M$}
	{
		$J = \argmax_j z_j$.\;
		$p_{i,J} = \min\{1,z_J\}$.\;
		$z_J = z_J - p_{i,J}$.\;
		\While {$\sum_j p_{i,j} <1$}
		{
			$K = \argmin_{k \in \{k|z_k>0\}} \|y_J - y_k\|$.\;
			$p_{i,K} = \min\{1-\sum_j p_{i,j}, z_K\}$.\;
			$z_K = z_K - p_{i,K}$.\;
		}
		$x_i = \sum_k p_{i,k}y_k$.\;
	}
\caption{The approximate multinomial resampler (AMR).\label{alg:AMR}}
\end{algorithm}
\end{table}

@@

\begin{figure}[htb]
\centering
\subfigure[Relative error in $\mathbb{E}(X)$]{\includegraphics[width=0.45\textwidth]{"figures/Resampler_EX"}}
\subfigure[Relative error in
$\mathbb{E}(X^2)$]{\includegraphics[width=0.45\textwidth]{"figures/Resampler_EX2"}}\\
\subfigure[Relative error in $\mathbb{E}(X^3)$]{\includegraphics[width=0.45\textwidth]{"figures/Resampler_EX3"}}
\subfigure[Cost of resamplers per significant figure]{\includegraphics[width=0.45\textwidth]{"figures/Resampler_speed"}}\\
\caption{Comparison of the performance between different resampling schemes. The example in Section~\ref{sec:problem 1} is implemented for this demonstration.}
\label{fig:AMR}
\end{figure}

To test the accuracy and speed of the three resamplers (ETPF,
bootstrap and AMR), we drew a sample of size $M$ from the proposal distribution
$\mathcal{N}(1,2)$. Importance weights were assigned, based on a
target distribution of $\mathcal{N}(2,3)$. The statistics of the
resampled outputs were compared with the original weighted samples. Figure \ref{fig:AMR} (a)-(c) show how the relative errors in the first
three moments of the samples changes with ensemble size $M$ for the three different
samplers. As expected, the AMR lies somewhere between the high
accuracy of the ETPF and the less accurate bootstrap
resampling. Note that only the error for the bootstrap multinomial
sampler is presented for the first moment since both the ETPF and the
AMR preserve the mean of the original weighted samples up to machine precision. Figure \ref{fig:AMR} (d) shows how the computational cost,
measured in seconds, scales with the ensemble size for the three
different methods. These results demonstrate that the AMR behaves how we wish, and
importantly ensures that exactly one sample of the output will lie in
each region with weights up to $\frac{1}{M}$ of the total.

@@

In this section we demonstrate convergence of the PAIS algorithm. We
also investigate the convergence properties of PAIS compared with
naively parallelised Metropolis-Hastings samplers. We assess the
performance of the PAIS algorithm using the relative $L^2$ error
defined in \eqref{eqn:L2_error}, as well as the relative error in the
first moment \eqref{eq:34567}.

In the one-dimensional examples which follow, we have optimised the
naively parallelised RWMH algorithm using the optimal acceptance rate
$\hat{\alpha} = 0.5$. This value differs from the theoretical
asymptotic value of 0.234 which applies in higher dimensions, but this
higher acceptance rate is commonly used for one dimensional Gaussian
posteriors~\cite{rosenthal2011optimal}. To find the optimal scaling
parameter we minimise the statistic
\[
	T_{\text{MH}}(\beta) = \left| \frac{N_{\text{acc}}(\beta)}{N_{\text{total}}} - \hat{\alpha} \right|,
\]
where $N_{\text{acc}}(\beta)$ is the number of accepted moves and
$N_{\text{total}}$ is the total number of samples produced. For the
PAIS algorithm, we maximise the effective sample size as
discussed in Section~\ref{sec:ess}.

\subsection{Sampling from a one dimensional Gaussian distribution}\label{sec:problem 1}

This first example shows how PAIS handles searching for the mode of a
Gaussian distribution when the initial state is a long way out in the
tails.

@@

Consider the simple case of a linear observation operator $\G(x) = x$,
where the prior on $x$ and the observational noise follow centred
Gaussian distributions. Then, following \eqref{eqn:like}, the Gaussian
posterior has the form
\begin{equation}\label{eqn:Gaussian posterior}
	\text{law}(\mu_D) = \pi(x|D) \propto \exp\left(-\frac{1}{2}\big\|x - D\big\|^2_{\sigma^2} - \frac{1}{2}\big\|x\big\|^2_{\tau^2}\right),
\end{equation}
where $\sigma^2$ and $\tau^2$ are the variances of the
observational noise and prior distributions respectively. In the
numerics which follow, we choose $\tau^2 =0.01$ and $\sigma^2 = 0.01$,
and we observe $x_\text{ref}=4$ noisily such that
\[
D = \mathcal{G}(x) + \eta \sim \mathcal{N}(\mathcal{G}(x_\text{ref}),\sigma^2) = \mathcal{N}(4,0.01).
\]
These values result in a posterior density in which the vast majority
of the density is out in the tails of the
prior distribution. The Kullback-Leibler (KL)
divergence, which gives us a measure of how different the prior and
posterior are, is $D_{KL}(\mu_D || \mu_0) = 4.67$ for this
problem. A KL divergence of zero indicates that two distributions are
identical almost everywhere.

\subsubsection{Numerical implementation}\label{sec:Implementation P1}

In each of the following simulations, we perform three tasks. First we
calculate the optimal value of $\beta$ by optimising the statistics
described in Section~\ref{sec:statistics}. We then run the algorithms
with optimal parameters to calculate and compare the convergence
rates. Finally, we implement the adaptive algorithms described in
Section~\ref{sec:adapt} and compare the convergence rates of these
algorithms with the nonadaptive algorithms.

{\bf (1) Finding the optimal parameters}: To find the optimal
parameters we choose 32 values of $\beta$ evenly spaced on a log scale
in the interval $[10^{-5}, 2]$. We run the PAIS-RW and RWMH algorithms
for one million iterations, each with an ensemble size of $M=50$. We took
32 repeats of both algorithms and then used the geometric means of the
sample statistics to find the optimal parameters.

{\bf (2) Measuring convergence of nonadaptive algorithms}: We run the
algorithms in Section~\ref{sec:problem 1} for one million iterations,
again with $M=50$. The simulations are repeated 32 times using the
optimal parameters found in (1). The performance of the algorithms is
judged by the convergence of the relative $L^2$ error statistic in
\eqref{eqn:L2_error}.

{\bf (3) Measuring convergence of adaptive algorithms}: We run the
adaptive algorithms under the same conditions as the nonadaptive
algorithms, and again use the relative $L^2$ error to compare
efficiency. The adaptive algorithms are initialised with $\beta^{(1)}$ = 1.

\subsubsection{Optimal values of $\beta$}\label{sec:Optimal values P1}

@@

Both adaptive algorithms are run with initial values of
$\beta=1$. Figure~\ref{fig:MH1 L2} shows that after an initial burn-in
period the APAIS-RW algorithm catches up to the PAIS-RW algorithm, and
by the end of the simulation window is matching its performance. The
ARWMH algorithm does not perform quite as well, this is possibly due
to the fact that $T_{\text{MH}}(\beta)$ is not smooth at the optimal
value making it difficult to minimise.

@@

Throughout this example, we use an ensemble size $M=50$, but it is
interesting to see how the PAIS algorithm scales when we increase the
ensemble size, and if there is some limit below which the algorithm
fails. We implement the problem in Section~\ref{sec:problem 1}, using
the RWMH and PAIS-RW algorithms with ensemble sizes in the interval $M
\in [1, 160]$.

\begin{figure}[h]
\begin{center}
\includegraphics[width=0.6\textwidth]{"figures/PAIS_saving"}
\caption{Ratio of PAIS-RW samples required to reach the same tolerance as the RWMH algorithm.}
\label{fig:PAIS_saving}
\end{center}
\end{figure}

Figure~\ref{fig:PAIS_saving} was produced using the method of finding
optimal $\beta$ described in Section~\ref{sec:Implementation P1} (1),
then running 32 repeats at each ensemble size. The convergence rates
are then found by regressing through the data. The graph is still very
noisy but demonstrates that increasing the ensemble size continues to
reduce the number of iterations required in comparison with naively
parallelised MH. The decreasing trend shows superlinear improvement of
PAIS with respect to ensemble size, in terms of the number of
iterations required, which is a demonstration of our belief that
parallelism of MCMC should give us added value over and above that
provided by naive parallelism. This decrease is due to the increasing
effective sample size shown in Figure~\ref{fig:neff-M} (b).

\subsection{Sampling from Bimodal Distributions}\label{sec:bimodal}

In this second example we investigate the behaviour of the PAIS algorithm
when applied to bimodal problems. MH methods can
struggle with multimodal problems, particularly where switches between
the modes are rare, resulting in incorrectly proportioned modes in the
histograms. This example demonstrates that the PAIS algorithm redistributes 
chains to new modes as they are found. This means that we expect the number of chains in a mode to be
approximately proportional to the probability density in that mode. As a result, reconstructed posteriors with disproportional modes, as is familiar with the MH algorithms, are not produced.

\subsubsection{Target Distribution} \label{sec:tar}

We look at an `easy' problem, $B_1$, which has a KL divergence of
0.880, and a `harder' problem, $B_2$, which has a KL divergence of
3.647. Problem $B_1$ has two modes which are not too far apart. In
$B_2$ we increase the distance between the two modes which has the
effect of increasing the expected number of iterations that it takes
for a MCMC chain to jump between modes. These posteriors are shown in Figure~\ref{fig:problem 3 posteriors}.

The following setup is the same for both problems. We consider a
non-linear observation operator $\G(x) = x^2$, and assign the prior
$x \sim \mu_0 = \N(0, \tau^2=0.25)$. We assume that a noisy reading,
$D$, is taken according to $D = \G(x_\text{ref}) + \varepsilon$, where
$\varepsilon \sim \mu_\varepsilon = \N(0, \sigma^2 = 0.1)$. This results
in the non-Gaussian posterior
\[
	\pi(x|D) \propto \exp\left(-\frac{1}{2\sigma^2}\|x^2 - D\|^2 - \frac{1}{2\tau^2}\|x\|^2\right).
\]
To create the `easy' problem we say that the true value of
$\G(x_\text{ref}) = 0.75$, and the `hard' problem is generated using
$\G(x_\text{ref}) = 2$. In the numerics which follow we draw noise from
$\mu_\varepsilon$ to generate our data point.

\begin{figure}[htpb]
\begin{center}
\includegraphics[width=\textwidth]{"figures/posteriors3"}
\caption{Posterior distributions for problems $B_1$ and $B_2$ as
  described in Section \ref{sec:tar}.}
\label{fig:problem 3 posteriors}
\end{center}
\end{figure}

\subsubsection{Calculating values of Optimal $\beta^*$}\label{sec:BM1_opt_beta}

Calculating the optimal values of the scaling parameters for this
problem is similar to the previous example; we check only the acceptance
rate to find the optimal values for RWMH and we use the effective
sample size to find the optimal values for PAIS-RW.
Table~\ref{table:BM_opt_beta} gives the optimal values of $\beta$ for both
problems. The subscript on $\beta$ refers to the criterion which has been optimised.

\begin{table}[!htb]
    \begin{minipage}{.5\linewidth}
      \centering
        \begin{tabular}{|l|r|r|}
	\hline
	Algorithm							& $\beta^*_{\text{acc}}$	& $\beta^*_{\text{eff}}$ \\ \hline
	RWMH								& 4.8e-1					& - \\
	PAIS-RW								& -						& 1.0e-1\\
	\hline
	\end{tabular}
    \end{minipage}%
    \begin{minipage}{.5\linewidth}
      \centering
        \begin{tabular}{|l|r|r|r|}
	\hline
	Algorithm							& $\beta^*_{\text{acc}}$	& $\beta^*_{\text{eff}}$	& $\beta^*_{\text{L2}}$ \\ \hline
	RWMH								& 2.3e-1					& - 						& 9.3e-1\\
	PAIS-RW								& -						& 5.1e-2 					& 1.3e-1\\
	\hline
	\end{tabular}
    \end{minipage}
	\vspace{1mm}
	\caption{Optimal values of $\beta$ for $B_1$ (left) and $B_2$ (right).}
	\label{table:BM_opt_beta}
\end{table}

It is relatively simple to find optimal scaling parameters for problem $B_1$.
These values are given in Table~\ref{table:BM_opt_beta} (left). However
problem $B_2$ is much harder as transitions between the modes are extremely
unlikely for the standard RWMH algorithm. This means that we need to consider the
convergence on two levels; we should consider the algorithm's ability
to find both the modes, and also whether it can sample them in the correct proportions.

To get correctly proportioned modes with the RWMH algorithm it is
important that the chains can transition between the modes frequently, which
means that $\beta$ must be large. However, this leads to a lower
acceptance rate, and so we sacrifice convergence locally. For this reason, the RWMH algorithm is very slow to
converge for problems of this type.

We can achieve these two regimes in RWMH by tuning $\beta$ using the acceptance rate for local convergence, and by $L^2$ error for global convergence. Similarly in PAIS-RW we can use the effective sample size for local convergence, and the $L^2$ error for global convergence.

From Table~\ref{table:BM_opt_beta} (right) we see that there is a large difference between the optimal value of $\beta$ for each regime in RWMH, and so will result in inefficient sampling. The PAIS-RW algorithm manages to sample the local detail and the large scale behaviour with similar values of $\beta^*$; a clear advantage to using this algorithm for this problem.

@@

\begin{figure}[htb]
\centering
\subfigure[Relative $L^2$ error.]{\includegraphics[width=0.45\textwidth]{"figures/BM1_L2"}}
\subfigure[Absolute error in first moment.]{\includegraphics[width=0.45\textwidth]{"figures/BM1_moments"}}
\caption{Convergence of the PAIS-RW and RWMH algorithms for Problem $B_1$. Set up described in Section~\ref{sec:Implementation P1}. Resampling is performed using the ETPF.}
\label{fig:BM1_L2}
\end{figure}

As in the Gaussian example we see a significant speed up with the
PAIS-RW algorithm for problem $B_1$. Figure~\ref{fig:BM1_L2} shows the adaptive and nonadaptive convergence rates. We can see that the adaptive algorithms compare closely with the
respective nonadaptive algorithms and the improvement PAIS offers remains significant.

\begin{figure}[htb]
\centering
\subfigure[Relative $L^2$ error using the optimal scaling parameters.]{\includegraphics[width=0.45\textwidth]{"figures/BM2_L2"}}
\subfigure[Comparison of the relative $L^2$ error of the algorithms optimised by $L^2$ minimisation with locally optimised PAIS-RW with a scout chain.]{\includegraphics[width=0.45\textwidth]{"figures/BM2_scout"}}
\caption{Convergence of the PAIS-RW and RWMH algorithms for Problem $B_2$. Set up described in Section~\ref{sec:Implementation P1}. Resampling is performed using the ETPF.}
\label{fig:BM2_L2}
\end{figure}

For $B_2$ the algorithms are run with the global optimal value of $\beta^*$, and with the local optimal value of $\beta^*$. Figure~\ref{fig:BM2_L2} (a) shows that the algorithms using the globally optimal $\beta^*$ convergence at the desired rate, whereas the algorithms optimised using the acceptance rate and effective sample size initially converge faster but at some point forget the location of one of the modes, causing the convergence to flatten out.

Using the PAIS algorithm we can minimise the impact of forgotten modes by constantly allowing the algorithm to search them out. Since we have parallel chains, we can run PAIS-RW with the majority of chains using the locally optimal value of $\beta$, and one or two chains with a larger scaling parameter. These chains with larger scaling parameters act both as `scouts' for new modes, and act to aid in the over-dispersal of the proposal distribution. Figure~\ref{fig:BM2_L2} (b) shows the results of using 49 chains with the local optimal scaling parameter, and one chain with ten times the local optimal scaling parameter. We see that modes are not forgotten and the algorithm converges with the improvement we see from PAIS-RW in the other problems. Other methods of mode searches are described in \cite{lan2013wormhole}.

The adaptive algorithm can just as easily be applied to the PAIS algorithm with the `scout' chains described in the previous paragraph. Since we need two equally sized sub-ensembles, we will use two groups of 24 ensembles with the same proposal distributions, and each group will also have a `scout' chain with a scaling parameter ten times that of the rest of the group.

>>>>

\documentclass[final]{siamltex}
%test change
\usepackage{cite}
\usepackage{graphicx,bbm,pstricks,soul}
\usepackage{pifont}
\usepackage{bbm,algorithmic,mdframed,placeins,multirow,booktabs,subfigure}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}

@@

\def\leftrightarrowfill@{%
 \arrowfill@\leftarrow\relbar\rightarrow%
 }
\newcommand*{\centerfloat}{%
  \parindent \z@
  \leftskip \z@ \@plus 1fil \@minus \textwidth
  \rightskip\leftskip
  \parfillskip \z@skip}
\makeatother
\makeatother

@@

\begin{table}[!ht]
\centering
\begin{algorithm}[H]
\DontPrintSemicolon
\BlankLine
	Set $\X^{(0)} = \X_0 = [x_1^{(0)},x_2^{(0)},\ldots,x_M^{(0)}]^\top$.\;
	\For {$i = 1, \dots, N$}
	{
		Sample $\Y^{(i)} = [y_1^{(i)},y_2^{(i)},\ldots,y_M^{(i)}]^\top, \quad y_j^{(i)} \sim
\nu(\cdot;x_j^{(i-1)})$.\label{algline:PAIS_propose}\;
		Calculate $\W^{(i)} = [w_1^{(i)},w_2^{(i)},\ldots,w_M^{(i)}]^\top,$ \quad $w^{(i)}_j =
\frac{\pi(y_j^{(i)})}{\chi(y_j^{(i)};\X^{(i-1)})}$, where
		\[
			\chi(\cdot;\X^{(i-1)}) = \frac{1}{M}\sum_{j=1}^M \nu(\cdot;x_j^{(i-1)}).
		\]

@@

\section{Automated tuning of algorithm parameters}\label{Sec:adapt}

@@

\begin{table}[!ht]
\centering
\begin{algorithm}[H]
\DontPrintSemicolon
\BlankLine
	Define update times $\{n_k\}_k$.\;
	\For{$n=1,\dots,N$}{
		Complete steps \ref{algline:PAIS_propose}-\ref{algline:PAIS_resample} of Algorithm~\ref{alg:PAIS}.\;
		\If {$n \in \{n_k\}$}
		{
			Divide ensemble into two halves. Use these halves to estimate the gradient in $\neff$ at $\beta_k$.\;
			Update $\beta_k$ using gradient ascent,
				\[
					\beta_{k+1} = \beta_k + \gamma\nabla\neff.
				\]
			\label{algline:gradient_ascent}
		}
	}
\caption{Adaptively tuned PAIS algorithm.\label{alg:adaptPAIS}}
\end{algorithm}
\end{table}
Our adaptive algorithm is given in Algorithm~\ref{alg:adaptPAIS}. We choose update times which, as suggested in Section~\ref{sec:ess} allow for a reasonable estimate of the effective sample size, but do not waste too many iterations. In Step~\ref{algline:gradient_ascent}, the gradient ascent parameter $\gamma$ may decrease over time, e.g. as a function of $n_k$.

@@

\begin{table}[!ht]
\centering
\begin{algorithm}[H]
\DontPrintSemicolon
\BlankLine
	$\b{z} = M\bar{\b{w}}$.\;
	\For {$i = 1,\dots, M$}
	{
		$J = \argmax_j z_j$.\;
		$p_{i,J} = \min\{1,z_J\}$.\;
		$z_J = z_J - p_{i,J}$.\;
		\While {$\sum_j p_{i,j} <1$}
		{
			$K = \argmin_{k \in \{k|z_k>0\}} \|y_J - y_k\|$.\;
			$p_{i,K} = \min\{1-\sum_j p_{i,j}, z_K\}$.\;
			$z_K = z_K - p_{i,K}$.\;
		}
		$x_i = \sum_k p_{i,k}y_k$.\;
	}
\caption{The approximate multinomial resampler (AMR).\label{alg:AMR}}
\end{algorithm}
\end{table}

@@

\begin{figure}[htb]
\centering
\subfigure[Relative error in $\mathbb{E}(X)$]{\includegraphics[width=0.45\textwidth]{"figures/Resampler_EX"}}
\subfigure[Relative error in
$\mathbb{E}(X^2)$]{\includegraphics[width=0.45\textwidth]{"figures/Resampler_EX2"}}\\
\subfigure[Relative error in $\mathbb{E}(X^3)$]{\includegraphics[width=0.45\textwidth]{"figures/Resampler_EX3"}}
\subfigure[\hl{Cost of resamplers per significant figure}]{\includegraphics[width=0.45\textwidth]{"figures/Resampler_speed"}}\\
\caption{Comparison of the performance between different resampling schemes. The example in Section~\ref{sec:problem 1} is implemented for this demonstration.}
\label{fig:AMR}
\end{figure}

To test the accuracy and speed of the three resamplers (ETPF,
bootstrap and AMR), we drew a sample of size $M$ from the proposal distribution
$\mathcal{N}(1,2)$. Importance weights were assigned, based on a
target distribution of $\mathcal{N}(2,3)$. The statistics of the
resampled outputs were compared with the original weighted samples. Figure \ref{fig:AMR} (a)-(c) show how the relative errors in the first
three moments of the samples changes with ensemble size $M$ for the three different
samplers. As expected, the AMR lies somewhere between the high
accuracy of the ETPF and the less accurate bootstrap
resampling. Note that only the error for the bootstrap multinomial
sampler is presented for the first moment since both the ETPF and the
AMR preserve the mean of the original weighted samples up to machine precision. Figure \ref{fig:AMR} \hl{(d) shows how the computational cost,
measured in seconds, scales with the ensemble size for the three
different methods}. These results demonstrate that the AMR behaves how we wish, and
importantly ensures that exactly one sample of the output will lie in
each region with weights up to $\frac{1}{M}$ of the total.

@@

In this section we demonstrate convergence of the PAIS algorithm. We
also investigate the convergence properties of PAIS compared with
naively parallelised Metropolis-Hastings samplers. We assess the
performance of the PAIS algorithm using the relative $L^2$ error
defined in Equation~\eqref{eqn:L2_error}, as well as the relative error in the
first moment (Equation~\eqref{eq:34567}).

\subsection{Numerical implementation}
\label{sec:Implementation P1}

For each example, we perform the following three tasks.

\begin{enumerate}
	\item {\bf Finding the optimal scaling parameters}: We chose 32 values of $\beta$ evenly spaced on a log scale in the interval $[10^{-5}, 2]$. We ran the PAIS and MH algorithms for one million iterations, each with an ensemble size of $M=50$. We took 32 repeats and used the geometric means of the sample statistics in Section~\ref{sec:statistics} to find the optimal scaling parameter.
	\item {\bf Measuring convergence of nonadaptive algorithms}: We ran the PAIS and MH algorithms, using the scaling parameters found in Step 1, for one million iterations, again with $M=50$. The simulations are repeated 32 times. The performance is evaluated using the relative $L^2$ error defined in Equation~\eqref{eqn:L2_error}.
	\item {\bf Measuring convergence of adaptive algorithms}: We run the
adaptive algorithms under the same conditions as the nonadaptive
algorithms, and again use the relative $L^2$ error to compare
efficiency. The adaptive algorithms are initialised with $\beta_1$ = 1.
\end{enumerate}

\subsection{Sampling from a one dimensional Gaussian distribution}
\label{sec:problem 1}

This first example shows how PAIS handles searching for the mode of a
Gaussian distribution when the initial state is a long way out in the
tails. We optimise the
naively parallelised RWMH algorithm using the optimal acceptance rate
$\hat{\alpha} = 0.5$. This value differs from the theoretical
asymptotic value of 0.234 which applies in higher dimensions. This
higher acceptance rate is commonly used for one dimensional Gaussian
posteriors~\cite{rosenthal2011optimal}. To tune the
PAIS algorithm we maximise the effective sample size as
discussed in Section~\ref{sec:ess}.

@@

Consider the case of a linear observation operator which maps a one
dimensional parameter onto the observale space, $\G\colon x \mapsto
x$. We assign centred priors to the parameter $x$, as well as to the
observational noise. We choose the true value of the parameter to be
$x_\text{ref} = 4$, then following Equations~\eqref{eqn:obs} and
\eqref{eqn:like},
\begin{equation}\label{eqn:Gaussian posterior}
	\text{law}(\mu_D) = \pi(x|D) \propto \exp\left(-\frac{1}{2}\big\|x
	 - D\big\|^2_{\sigma^2} - \frac{1}{2}\big\|x\big\|^2_{\tau^2}\right).
\end{equation}

This set-up results in a posterior density in which the vast majority
of the density is out in the tails of the
prior distribution. The Kullback-Leibler (KL)
divergence, which gives us a measure of how different the prior and
posterior are, is $D_{KL}(\mu_D || \mu_0) = 4.67$ for this
problem. A KL divergence of zero indicates that two distributions are
identical almost everywhere.

\subsubsection{Optimising the scaling parameter}\label{sec:Optimal values P1}

@@

Figure~\ref{fig:MH1 L2} shows that after an initial burn-in
period the APAIS-RW algorithm catches up to the PAIS-RW algorithm, and
by the end of the simulation window is matching its performance. The
ARWMH algorithm does not perform quite as well, this is possibly because
it is optimising a cost functional which is not smooth.

@@

In this and the next example we use an ensemble size $M=50$, however it is
interesting to see how the PAIS algorithm scales when we increase the
ensemble size, and if there is a minimum required ensemble size. We
implement the problem in Section~\ref{sec:problem 1}, using
the RWMH and PAIS-RW algorithms with ensemble sizes in the interval $M
\in [1, 160]$.

\begin{figure}[ht]
\begin{center}
\includegraphics[width=0.6\textwidth]{"figures/PAIS_saving"}
\caption{Ratio of PAIS-RW samples required to reach the same tolerance as the RWMH algorithm.}
\label{fig:PAIS_saving}
\end{center}
\end{figure}

Figure~\ref{fig:PAIS_saving} was produced using the method of finding
optimal $\beta$ described in Section~\ref{sec:Implementation P1} (1),
then running 32 repeats at each ensemble size

We repeat the process described in Section~\ref{sec:Implementation P1}
for each of the ensemble sizes we are interested in. The convergence rates
are then found by regressing through the data. The graph is still very
noisy but demonstrates that increasing the ensemble size continues to
reduce the number of iterations required in comparison with naively
parallelised MH. The decreasing trend shows superlinear improvement of
PAIS with respect to ensemble size, in terms of the number of
iterations required, which is a demonstration of our belief that
parallelism of MCMC should give us added value over and above that
provided by naive parallelism. This decrease is due to the increasing
effective sample size shown in Figure~\ref{fig:neff-M} (b).

\subsection{Sampling from Bimodal Distributions}
\label{sec:bimodal}

In this second example we investigate the behaviour of the PAIS
algorithm when applied to bimodal problems. MH methods can struggle
with multimodal problems, particularly where switches between the
modes are rare, resulting in incorrectly proportioned modes in the
histograms. This example demonstrates that the PAIS algorithm
redistributes particles to new modes as they are found. This means
that we expect the number of particles in a mode to be approximately
proportional to the probability density in that mode, resulting in
faster global convergence.

\subsubsection{Target distribution} \label{sec:tar}

\begin{figure}[htpb]
\begin{center}
\includegraphics[width=\textwidth]{"figures/BM2_posterior"}
\caption{Posterior distribution for the bimodal example in
Section~\ref{sec:tar}.}
\label{fig:BM2_posterior}
\end{center}
\end{figure}

In this example we look at a bimodal posterior distribution which has
a high energy barrier between the two modes. In the MH algorithm, the
expected number of iterations for a switch between modes is a large
proportion of the total number of iterations. This posterior is in
Figure~\ref{fig:BM2_posterior}.

We consider a non-linear observation operator $\G\colon x \mapsto
x^2$, and assign the prior $\mu_0 = \N(0, \tau^2=0.25)$. Again, we
assume the true parameter $x_\text{ref}=2$ is observed according to
Equation~\eqref{eqn:obs} with observational noise $\sigma^2 = 0.1$,
and the posterior is
\[
	\pi(x|D) \propto \exp\left(-\frac{1}{2\sigma^2}\|x^2 - D\|^2
		 - \frac{1}{2\tau^2}\|x\|^2\right).
\]

\subsubsection{Calculating values of the optimal scaling parameter}
\label{sec:BM2_opt_beta}

Calculating the optimal values of the scaling parameters for this
example is similar to the previous example. These values are given in
Table~\ref{tab:BM2_opt_delta}. The subscript on $\beta$ refers to the
criterion which has been optimised.

\begin{table}[!htb]
      \centering
        \begin{tabular}{|l|r|r|r|}
	\hline
	Algorithm							& $\beta^*_{\text{acc}}$	& $\beta^*_{\text{eff}}$	& $\beta^*_{\text{L2}}$ \\ \hline
	RWMH								& 2.3e-1					& - 						& 9.3e-1\\
	PAIS-RW								& -						& 5.1e-2 					& 1.3e-1\\
	\hline
	\end{tabular}
	\vspace{2mm}
	\caption{Optimal scaling parameters for the example in Section~\ref{sec:bimodal}.}
	\label{tab:BM2_opt_delta}
\end{table}

Since transitions between the modes are extremely unlikely, we
consider the convergence on two levels; convergence to equally
proportioned modes, and convergence to smooth histograms.
To get correctly proportioned modes with the RWMH algorithm it is
important that the chains can transition between the modes frequently,
which requires a large proposal variance. However, this leads to a
lower acceptance rate, and so we sacrifice convergence locally. For
this reason, the RWMH algorithm is very slow to converge for problems
of this type.

We can achieve these two regimes in RWMH by tuning $\beta$ using
different statistics; the acceptance rate for local convergence, and
the $L^2$ error for global convergence. Similarly in PAIS-RW we can
use the effective sample size for local convergence, and the $L^2$
error for global convergence.

@@

\begin{figure}[htb]
\centering
\subfigure[Relative $L^2$ error using the optimal scaling parameters.]
{\includegraphics[width=0.45\textwidth]{"figures/BM2_L2"}}
\subfigure[Convergence of locally optimised PAIS-RW with scout chain,
and globally optimised PAIS-RW and RWMH.]
{\includegraphics[width=0.45\textwidth]{"figures/BM2_scout"}}
\caption{Convergence of the PAIS-RW and RWMH algorithms for the
bimodal example discussed in Section~\ref{sec:bimodal}. Set up
described in Section~\ref{sec:Implementation P1}. Resampling is
performed using the ETPF.}
\label{fig:BM2_L2}
\end{figure}

The MH and PAIS algorithms are run with both the globally and locally
optimal scaling parameters. Figure~\ref{fig:BM2_L2} (a) shows that the
globally optimal $\beta^*$ leads to normal convergence rates,
$\mathcal{O}(n^{-1/2})$, whereas algorithms using locally optimal
scaling parameters initially converge faster but do not sample
proportionally for the whole simulation.

Using the PAIS algorithm we can ensure that we do not lose modes as
simulation progresses. Since we have an ensemble, we can sacrifice
locally optimal sampling for one or two ensemble members in favour of
a larger proposal variance. These chains with larger scaling
parameters act both as `scouts' for new modes, and act to aid in the
over-dispersal of the proposal distribution. Figure~\ref{fig:BM2_L2}
(b) shows the results of using 49 chains with the local optimal
scaling parameter, and one chain with ten times the local optimal
scaling parameter. We see that modes are not forgotten and the
algorithm converges with the improvement we see from PAIS-RW in the
Gaussian example.

The adaptive algorithm can just as easily be applied to the PAIS
algorithm with the `scout' chains described in the previous paragraph.
Since we need two equally sized sub-ensembles, we will use two groups
of 25 ensemble members. Each group has 24 ensemble members tuned for
locally optimal convergence and one `scout' chain with a scaling
parameter ten times that of the rest of the group.

\begin{figure}[htb]
\begin{center}
\includegraphics[width=0.6\textwidth]{"figures/BM2_AL2"}
\caption{Convergence of modified PAIS and RWMH algorithms, comparing
the globally optimised nonadaptive algorithms with the locally
optimised adaptive algorithms. The setup is as described in
Section~\ref{sec:Implementation P1} (3). Resampling is performed using
the ETPF.}
\label{fig:BM2_AL2}
\end{center}
\end{figure}

Comparing the convergence of the adaptive algorithms against the
nonadaptive algorithms in Figure~\ref{fig:BM2_AL2} shows that the
algorithms behave as expected. The adaptive RWMH algorithm tuned using
the acceptance rate converges at the same rate as the $L^2$ optimised
algorithm until the scaling parameter gets too small, and therefore
switches between the modes are rare, and the relative heights of the
modes are decided by the arbitrary proportion of chains which are in
each mode at this point. The adaptive PAIS-RW with scout chains tuned
to the effective sample size converges at about the same rate as the
locally optimised nonadaptive algorithm also with scouts.

\subsubsection{Calculating the speed up in convergence}
\label{sec:calc_saving}
